\texttt{TREE}: A package for modelling plant TRait Ecology and Evolution

### 2015-05-28 ms.tex
update schematic
For #14
Build from parts via tikz
Change fonts
Feedback from James

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Population dynamics in forests are strongly size-structured:
  larger plants shade smaller plants while also expending
  proportionately more energy on building and maintaining woody stems.
  Although the importance of size-
  structure for demography is widely recognised, many mechanistic models
  either omit it entirely, or include only coarse approximations.
\item
  Here we introduce the \texttt{TREE} package (TRait Ecology and Evolution), an
  extensible framework for modelling plant demography across the entire
  life cycle via coupled differential equations. At its core, \texttt{TREE} is an
  individual-based mechanistic model where plant function is mediated by
  traits. Individuals from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
  disturbance regime. Dynamics within patches of competing plants are
  resolved using novel extensions of the Escalator Boxcar Train
  technique. Combined effects of trait-, size- and patch-structured
  dynamics are integrated into population level estimates of
  reproductive fitness.
\item
  \texttt{TREE} is presented as an open source \texttt{R} package and is
  available at
  \href{https://github.com/traitecoevo/tree}{github.com/traitecoevo/tree}.
  While accessed from R, the core routines in \texttt{TREE} are written in C++.
  The package provides for alternative physiologies and for
  hyper-parameterisation on the basis of plant functional traits. A
  detailed test suite is provided to ensure accuracy.
\item
  We provide worked examples illustrating \texttt{TREE}'s use to study the
  influence of functional traits on growth of individual plants, of
  whole patches and on assembly of ecological communities.
\end{enumerate}

@@

Plant growth and demography are fundamentally size- and trait- structured,
influencing dynamics over time-scales ranging from instantaneous physiological
effects to long-term evolutionary outcomes \citep{Harper-1977, Westoby-2002}. As
an individual increases its leaf area, its potential to generate
photosynthetic also rises. Yet for larger individuals,  increasing
fractions of photosynthetic income must also be diverted towards
building and maintaining support tissues rather than leaves
\citep{Givnish-1988, Enquist-2007}. A larger fraction of income is also
allocated to reproduction \citep{Thomas-2011}. Consequently at
population level, rates of growth and mortality change with size
\citep{Muller-2006, Thomas-2011, Ruger-2011, Wright-2010}. The traits of
a species also interact with size to shape its demography. Leaf
construction costs and wood density affect growth rates
\citep{Wright-2010}. Other traits such as seed size and height at
maturation determine start and end points of ontogenetic trajectories
\citep{Westoby-2002}. Over longer time-frames, natural selection favours
traits making plants more competitive such as faster growth, larger
seeds, taller height \citep{Falster-2003}. Selective forces thus
potentially amplify size-related effects on demography. Accounting for
the effects of size and traits in models therefore seems like a key step
in developing theory for the structure and dynamics of the biosphere
\citep{Purves-2008, Moorcroft-2001, Falster-2003}, in particular notable
phenomena such as successional dynamics, self thinning, trait evolution
and species coexistence.

Although the importance of size-structure for vegetation function and
demography has been widely recognised for a long time
\citep{Harper-1977, Hara-1984, Shugart-1980, Huston-1987, Moorcroft-2001,
Kohyama-1993, Enquist-2007, Pacala-1996, Coomes-2007},
most large-scale mechanistic models of plant growth either omit it
entirely, or include only coarse approximations
\citep{Cramer-2001, Dekauwe-2014, Kelley-2013, Sitch-2003}. The likely
reason is the computational
challenge of modelling size-structured dynamics. During the 1980s and
early 1990s, size-structured ``gap'' models were widely used
\citep[e.g.][]{Huston-1987, Shugart-1980}. These gap models were
parameterised in terms of growth and demography, and consequently did a
poor job of simulating carbon fluxes. The next generation of models
focussed more on physiology and, presumably for computational reasons,
sacrificed details of demography in order to achieve global coverage
\citep{Haxeltine-1996}. This history has brought us to the
current perplexing situation where all but a couple
\citep{Moorcroft-2001, Smith-2014} of the leading vegetation
models lack size-structure. These models
are focused on modelling fluxes carbon and water, but say little about
demographic behaviour, and consequently are also unable to account for
any non-linear demographic feedbacks that may arise.

Interestingly, most models used in theoretical ecology to explore
questions about species coexistence also lack size-structured dynamics
\citep[e.g.][]{Calcagno-2006, Geritz-1995, Leimar-2013, Levin-1974,MacArthur-1967,Tilman-1985}.
It is common to assume competitive interactions are influenced by
size-related traits such as adult and seed size, but detailed
size-structured demography is rarely considered, at least for plant
models \cite[for animal examples, see][]{Deroos-1988, Deroos-1992}. Thus it has
been difficult to connect mechanisms described in these simplistic
models with field data and real traits, which are almost always strongly
size-structured.

In this note, we describe the \texttt{TREE} package for R
\citep{R-2015}, a mechanistic framework for studying the effects of size
structure and trait variation on the demography of individual plants, of
patches of competing plants, and of meta-populations structured by a
prevailing disturbance regime. The package came about as a way to extend
previous work investigating the effect of traits on vegetation
properties \citep{Falster-2011} and the mechanisms facilitating
coexistence of trait mixtures \citep{Falster-2015}. These studies
provided the foundation methods needed; here we provide robust and
extensible implementations. We describe the general approach of the
package and then a series of use cases, focussing at different levels of
ecological organisation. Each section provides a short description of
the required methods, design considerations, and then some worked
examples illustrating potential use cases and design features. Detailed
technical documents are provided as supplementary information, and updated
versions of technical documents are also available within the \texttt{TREE}
package itself. The use cases highlighted reflect our own interests in
understanding trait-based demography and community assembly. Yet we
expect the methods provided to be useful in other contexts. For example,
prior to \texttt{TREE} there was no \texttt{R} package that allowed one to simply grow a
plant over time, based on well-understood physiology. Surely this is one
of the simplest demographic actions one might hope for, with diverse
potential applications in ecology, evolution, forestry, agriculture, and
vegetation modelling. We hope the mechanistic platform provided here
will complement the increasingly sophisticated statistical methods now
available to analyse size-structured data \citep[e.g.][]{Metcalf-2013}.

@@

\texttt{TREE} is a mechanistic model, meaning the dynamics of the system arise
from rules specifying on how individuals grow and interact. In ecology,
there is a long history of using simple deterministic models
to understand biological phenomena, such as Lotka-Volterra population dynamics
\citep{MacArthur-1967, Leimar-2013} .
\texttt{TREE} was developed with this style of analysis in mind, while also
allowing for a richer set of ecological dynamics than is possible in
population models lacking size structure. The package implements methods
for physiological, population, and adaptive dynamics (Fig.
\ref{fig:schematic}) following methods described by \citet{Falster-2011}
and \citet{Falster-2015}. The core rules in \texttt{TREE} are about the
short-term physiological functioning of an individual plant and how
these are influenced by its traits, size and light environment. Dynamics
at higher levels of organisation then arise as emergent properties,
driven by growth physiology, competition for light and disturbance (Fig.
\ref{fig:schematic}). Demographic phenomena can be studied at the three
levels of individual plants, stands of competing plants, and entire
metapopulations.

@@

The core of \texttt{TREE} is a model for a plant's physiological strategy as
specified by its traits (Fig. \ref{fig:schematic}a). This sub-model
estimates rate of biomass production for a plant, given its size, light
environment, and the supplied physiological constants. Assimilation is
estimated from total leaf area and the light distribution across the
plant's canopy. Costs of tissue respiration and turnover are then
subtracted. The remaining biomass is then allocated between growth and
reproduction. The strategy model thus includes a list of physiological
rules and associated parameters.

The core job of the physiological model is to take size, light
environment and parameters as inputs, and to return growth rate,
mortality and fecundity as outputs, in other words to move from
individual growth to demography. However additional quantities are also
available as output, such as total assimilation and the respiration,
turnover, and allocation rates for different tissues.

The default physiological model used in \texttt{TREE} largely reflects that
presented in \citet{Falster-2011}, but with extensions allowing for diameter
growth also to be estimated. To properly model trait variation,
different parameters in the model are linked via assumed tradeoffs. For
example, the trait LMA is used to estimate the rate of leaf turnover,
based on widely observed scaling relationship from \citet{Wright-2004}.

@@

With above functionality, \texttt{TREE} can be used to estimate essential
physiological rates for individual plants (Fig. \ref{fig:plant}). The
function \texttt{grow\_plant\_to\_size} takes a given strategy and light
environment and grows the plant, producing a trajectory of size over
time (Fig. \ref{fig:plant}a). This is achieved by integrating an
Ordinary Differential Equation (ODE) of size-dependent growth rate (see
Appendix \ref{sec:demography} for more
details). As the plant grows, allocation to different types varies, and
the composition of the plant changes to include more stem and less
leaf (Fig. \ref{fig:plant}b). This in turn affects maintenance and
turnover costs and the growth rate of plants (Fig. \ref{fig:plant}c).
This shift in allocation alone generates a distinctive hump-shaped pattern of absolute
growth with size \citep{King-2011}, and equally the widely regarded
decline in relative growth rate with size from birth onwards
\citep{Enquist-2007}.

By varying the light environment and measuring growth rate, the whole
plant light compensation point (WPLCP) can be computed (Fig.
\ref{fig:plant}d). The WPLCP is increasingly regarded as the most useful
measure of a strategy's shade tolerance
\citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, WPLCP
increases with plant size, due to increased costs of building and
maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, WPLCP
decreases with LMA, because high LMA species have lower leaf turnover
\citep{Baltzer-2007, Lusk-2013}.

@@

When modelling a patch, we are interested in modelling changes in the
size-density-distribution \(n(h,a)\) over time, i.e.~the density of
individuals with height \(h\) within a patch of age \(a\). We assume
that patches are large, and are vertically but not horizontally
structured. Thus, shading in \texttt{TREE} is modelled as a mean field with
respect to spatial layout. Similar assumptions are found in many models
simulating size-structured dynamics
\citep{Moorcroft-2001, Huston-1987, Smith-2014}. Under these
assumptions, the dynamics of \(n\) behave deterministically and can be
approximated via a Partial Differential Equation (PDE) (see Appendix
\ref{sec:demography}). The same PDE has been shown to capture average behaviour
across a large number of small patches \citep{Moorcroft-2001}. Ideally
one would also consider full spatial interactions within patches,
however such models are very computationally demanding, and also
impossible to solve in a deterministic fashion \citep{Pacala-1996}.
Moreover, it remains unclear whether resolving spatial details within a
patch would provide further ecological detail, in part because the
process of competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

Our approach for modelling solving size-structured population dynamics
is based on the Escalator Boxcar Train technique (EBT)
\citep{Deroos-1997, Deroos-1992, Deroos-1988}. The EBT solves the PDE
describing development of \(n\) by approximating the density function
with a collection of cohorts spanning the size spectrum. Following a
disturbance, a series of cohorts are introduced into each patch. These
cohorts are then transported along the characteristic equations of the
PDE. Biologically these are the growth trajectories of individuals,
conditioned on them surviving (e.g. Fig. \ref{fig:patch}a). Meanwhile,
growth and mortality combine to alter the density of individuals along
the trajectories (Fig. \ref{fig:patch}b).

@@

The extension for handling strong size-asymmetric competition involves
adaptive refining of time points at which new cohorts are introduced
into the population. In the original EBT, cohorts were introduced at a
fixed time interval. Under competition, the growth trajectories of
individuals born at similar times can diverge substantially over time
(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography}).
This is can become a problem later during patch
development because much of the population can be located between
widely-spaced cohorts, delivering low numerical precision. To maintain
numerical precision we therefore use an iterative algorithm to adaptively refine
the cohort spacing until the trajectories of adjacent cohorts remain
adequately resolved throughout development of the patch (see
Appendix \ref{sec:demography} for details). This refinement causes an uneven
distribution of cohort introduction times across patch age (Fig.
\ref{fig:patch}a).

@@

First, \texttt{TREE} can simulate patterns of patch development after
disturbance, including self-thinning behaviours and successional
replacement. Whereas previous approaches for modelling self-thinning
were limited to a single species
\citep[e.g.][]{Barnes-2004, Coomes-2007}, \texttt{TREE} accommodates multiple
different species with different traits (Fig. \ref{fig:schematic}b; Fig.
\ref{fig:patch}).

@@

Having simulated a single patch, it is straightforward to scale up from
a single patch to an entire metapopulation, at least for the situation
of metapopulations at an equilibrium mix of patch ages and where
disturbances remove all established vegetation. The frequency-density
\(p(a)\) of patches age \(a\) in the landscape is needed. Assuming there
are a large number of patches means the dynamics of \(p\) behave
deterministically and can be approximated via a second PDE
\citep{Vonfoerster-1959, Mckendrick-1926} (see Appendix
\ref{sec:demography} for
details). Scaling from patch to metapopulation is then achieved by
weighting the temporal dynamics of a single patch by \(p(a)\), the
relative abundance of patches age \(a\) in the metapopulation.

The main numerical challenge in stepping from single patch
to metapopulation is to identify the equilibrium seed rain of the
metapopulation. For a given input seed rain, the model produces an
output seed rain. The equilibrium seed rain is the vector of seed rain
where input seed rain equals output seed rain. Finding this equilibrium
requires finding a root of a multidimensional function, subject to the
constraints that this root is stable (e.g., the trivial equilibrium of
zero seed rain always exists, but is often not stable) (Appendix \ref{sec:fitness});

@@

Beyond its conceptual appeal, this idea of a dynamic equilibrium allows
us to close the demographic feedback loop and thereby create a
self-regulating system. The equilibrium seed rain for multi-species
communities is approached demographically, incorporating the combined
effects of the model's physiological rules, species traits, competitive
interactions and the prevailing disturbance regime on the
size-distributions of plants within a metapopulation of patches. Unlike
some vegetation models, there is no need to set the density of seeds or
seedlings, these arise as outputs from the model. Characteristics of the
vegetation such as total leaf area, biomass, productivity and allocation
to different plant parts arise as emergent properties (Fig.
\ref{fig:patch}c). Similarly, distributions of abundance, leaf area and
growth rate with respect to size emerge at both patch scale and scale of
metapopulation (Fig. \ref{fig:emergent}). Note in particular how a single
physiological model produces a variety of relationships, dependent on patch
age. These emergent phenomena might then be compared to the observed
distributions in large forest tracts \citep{Muller-2006}.

@@

With the extension to metapopulation, we are modelling demography across
the entire life cycle and are thus in a position to estimate the fitness
of different types. Invasion fitness quantifies the ability of a rare
individual with particular traits to establish in a community of
established residents at their equilibrium densities\citep{Metz-1992}.
Invasion fitness
is ideally calculated as the long-term per capita growth rate of the new
type; however, in structured metapopulations, a more convenient
indicator of per-capita growth rate is given by the logarithm of the
basic reproduction ratio \citep{Gyllenberg-2001, Metz-2001}. The basic
reproduction ratio \(R\) is simply the total seed output of the new
type, averaged across the metapopulation. The new type can invade when
\(R > 1\).

@@

The current parametrisation of the physiological model is based on
\citet{Falster-2011}. These values were drawn from available literature (see
Appendix \ref{sec:FFW16}); however for some parameters limited or no data were
readily available. Providing updated parameter sets is trivial (see Appendix
\ref{sec:parameters}). Ideally, parameters would be inferred through a combination of
independent estimates and constraining outputs from \texttt{TREE} against diverse data sources
\citep{Lebauer-2012, Keenan-2013}.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/schematic}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/plant.pdf}
\caption{Physiological dynamics for individual plants vary with size, trait
and light environment. (a) Growth trajectories for plants differing in light
environment and trait (LMA, low = 0.1, solid line; high =  0.267, dashed
line). Lower light environments slow trajectories and interact non-linearly
with traits. (b) Over time, the fraction of living tissue switches from leaf
towards sapwood, with high LMA species having relatively more mass in leaf
than low LMA species. (c) Growth rates peak around 5m high, but this position
varies with both trait and light level (note that this is the derivative of
panel (a)). (d) Growth rates vary with plant size (seedling and mature shown)
and with traits; growth rate differences are more pronounced in high light.
Whole plant light compensation points emerge from the model as the point where
growth rate is zero (x intercept).  The light level used in panels (a) and (c)
is indicated with dots.}
\label{fig:plant}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/patch.pdf}
\caption{Population dynamics for two species competing within a patch. Lines
in panels (a) and (b) show the growth trajectories for individuals on the
boundary of successive "cohorts" for each species. A single cohort includes
all individuals born within a given time period. Red lines are the low LMA
species from figure \ref{fig:plant}, blue lines are the high LMA species.  The
initial growth rate advantage of the low LMA species (see Figure
\ref{fig:plant} c) means it quickly over tops the high LMA species,
suppressing its growth.  Self-thinning leaves space for the high LMA species
to establish below the canopy of the fast species. Panel (a) shows all
trajectories equally weighted; the spacing is generated by an adaptive
refinement algorithm.  In panel (b) trajectories are shaded by the density of
individuals at given size and patch age.  (c) The leaf area index at ground
level for each species.  The dotted line is the total leaf area index (sum
across both species).}
\label{fig:patch}
\end{figure}

>>>>

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Population dynamics in forests are strongly size-structured:
  larger plants shade smaller plants while also expending
  proportionately more energy on building and maintaining woody stems.
  Although the importance of size-
  structure for demography is widely recognised, many mechanistic models
  either omit it entirely, or include only coarse approximations.
\item
  Here we introduce the \texttt{TREE} package (TRait Ecology and Evolution), an
  extensible framework for modelling plant demography across the entire
  life cycle and in structured metapopulations, via coupled differential equations.
  At its core, \texttt{TREE} is an
  individual-based mechanistic model where plant physiology and demography is mediated by
  traits. Individuals from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
  disturbance regime. Dynamics within patches of competing plants are
  resolved using novel extensions of the Escalator Boxcar Train
  technique. Combined effects of trait-, size- and patch-structured
  dynamics are integrated into population level estimates of
  reproductive fitness.
\item
  \texttt{TREE} is presented as an open source \texttt{R} package and is
  available at
  \href{https://github.com/traitecoevo/tree}{github.com/traitecoevo/tree}.
  While accessed from R, the core routines in \texttt{TREE} are written in C++.
  The package provides for alternative physiologies and for
  hyper-parameterisation on the basis of plant functional traits. A
  detailed test suite is provided to ensure correct behaviour of the code.
\item
  We provide worked examples illustrating \texttt{TREE}'s use to study the
  influence of functional traits on growth of individual plants, of
  whole patches and on assembly of ecological communities.
\end{enumerate}

@@

Plant growth and demography are fundamentally size- and trait- structured,
influencing dynamics over time-scales ranging from instantaneous physiological
effects to long-term evolutionary outcomes \citep{Harper-1977, Westoby-2002}. As
an individual increases its leaf area, its potential to generate
photosynthate also rises. Yet as individuals get larger, an increasing
fraction of photosynthate must be diverted towards
building and maintaining support tissues rather than leaves
\citep{Givnish-1988, Enquist-2007}, and reproduction \citep{Thomas-2011}. Consequently 
rates of growth and mortality change with individual size
\citep{Muller-2006, Thomas-2011, Ruger-2011, Wright-2010}. The traits of
a species also interact with size to shape its demography. Leaf
construction costs and wood density affect growth rates
\citep{Wright-2010}. Other traits such as seed size and height at
maturation determine start and end points of ontogenetic trajectories
\citep{Westoby-2002}. Over longer time-frames, natural selection favours
traits that provide a competitive advantage, such as faster growth, larger
seeds or taller height \citep{Falster-2003}. Consequently, selective forces may amplify size-related effects on demography. Accounting for the effects of size and traits in models is therefore fundamental to developing theory on the structure and dynamics of the biosphere
\citep{Purves-2008, Moorcroft-2001, Falster-2003}, in particular for notable
phenomena such as successional dynamics, self thinning, trait evolution
and species coexistence.

Although the importance of size-structure for vegetation function and
demography has been long been recognised
\citep{Harper-1977, Hara-1984, Shugart-1980, Huston-1987, Moorcroft-2001,
Kohyama-1993, Enquist-2007, Pacala-1996, Coomes-2007},
most large-scale mechanistic models of plant growth either omit it
entirely, or only include coarse approximations
\citep{Cramer-2001, Dekauwe-2014, Kelley-2013, Sitch-2003}. The likely
reason is the computational
challenge of modelling size-structured dynamics. During the 1980s and
early 1990s, size-structured ``gap'' models were widely used
\citep[e.g.][]{Huston-1987, Shugart-1980}. These gap models were
parameterised in terms of growth and demography, and consequently did a
poor job of simulating carbon fluxes \todo{ref}. The next generation of models
focussed more on physiology and, presumably for computational reasons,
sacrificed details of demography in order to achieve global coverage
\citep{Haxeltine-1996}. This history has brought us to the
current perplexing situation where all but a couple
\citep{Moorcroft-2001, Smith-2014} of the most popular vegetation
models lack size-structure. Instead, most models
are focused on modelling fluxes carbon and water, but say little about
demographic behaviour, and consequently are also unable to account for
any non-linear demographic feedbacks that may arise.

Interestingly, most models used in theoretical ecology to explore
questions about species coexistence also lack size-structured dynamics
\citep[e.g.][]{Calcagno-2006, Geritz-1995, Leimar-2013, Levin-1974,MacArthur-1967,Tilman-1985}.
It is common to assume competitive interactions are influenced by
size-related traits such as adult size and seed size, but detailed
size-structured demography is rarely considered, at least for plant
models \cite[for animal examples, see][]{Deroos-1988, Deroos-1992}. Consequently, it has
been difficult to connect mechanisms described in these simplistic
models with field data and real traits, which are almost always strongly
size-structured.

In this note, we describe the \texttt{TREE} package for R
\citep{R-2015}, a mechanistic framework for studying the effects of size
structure and trait variation on the demography of individual plants, of
patches of competing plants, and of meta-populations structured by a
prevailing disturbance regime. The package came about as a way to extend
previous work investigating the effect of traits on vegetation
properties \citep{Falster-2011} and the mechanisms facilitating
coexistence of trait mixtures \citep{Falster-2015}. These studies
provided the foundation methods needed \todo{needed for what.. its hinted above... maybe rearrange the sentences} here we provide robust and
extensible implementations. We describe the general approach of the
package and then a series of use cases, focussing at different levels of
ecological organisation. Each section provides a short description of
the required methods, design considerations, and then some worked
examples illustrating potential use cases and design features. Detailed
technical documents are provided as supplementary information, and updated
versions of technical documents are also available within the \texttt{TREE}
package itself. The use cases highlighted reflect our own interests in
understanding trait-based demography and community assembly. Yet we
expect the methods provided to be useful in other contexts. For example,
prior to \texttt{TREE} there was no \texttt{R} package that allowed one to simply grow a
plant over time, based on well-understood physiology. Surely this is one
of the simplest demographic actions one might hope for, with diverse
potential applications in ecology, evolution, forestry, agriculture, and
vegetation modelling. We hope the mechanistic platform provided here
will complement the increasingly sophisticated statistical methods now
available to analyse size-structured data \citep[e.g.][]{Metcalf-2013}.

@@

\texttt{TREE} is a mechanistic model, meaning the dynamics of the system arise
from rules specifying how individuals grow and interact. In ecology,
there is a long history of using simple deterministic models
to understand biological phenomena, such as Lotka-Volterra population dynamics
\citep{MacArthur-1967, Leimar-2013} .
\texttt{TREE} was developed with this style of analysis in mind, while also
allowing for a richer set of ecological dynamics than is possible in
population models lacking size structure. The package implements methods
for physiological, population, and adaptive dynamics (Fig.
\ref{fig:schematic}) following methods described by \citet{Falster-2011}
and \citet{Falster-2015}. The core rules in \texttt{TREE} are about the
short-term physiological functioning of an individual plant and how
these are influenced by its traits, size and light environment. Dynamics
at higher levels of organisation then arise as emergent properties,
driven by growth physiology, competition for light and disturbance (Fig.
\ref{fig:schematic}). Demographic phenomena can be studied at the three
levels: individual plants, stands of competing plants, and entire
metapopulations.

@@

The core of \texttt{TREE} is a model for a plant's physiological strategy as
specified by its traits (Fig. \ref{fig:schematic}a). This sub-model
estimates rate of biomass production for a plant, given its size, light
environment, and the supplied physiological constants. Gross photosynthetic 
income is
estimated from total leaf area and the light distribution across the
plant's canopy. Costs of tissue respiration and turnover are then
subtracted. The remaining biomass is then allocated between growth and
reproduction. The strategy model thus includes a list of physiological
rules and associated parameters.

The core job of the physiological model is to take size, light
environment and physiological parameters as inputs, and to return growth rate,
mortality and fecundity as outputs, in other words to move from
individual growth to demography. However additional quantities are also
available as output, such as total assimilation and the respiration,
turnover, and allocation rates for different tissues.

The default physiological model used in \texttt{TREE} largely reflects that
presented in \citet{Falster-2011}, but with extensions allowing for diameter
growth also to be estimated. To properly model trait variation,
different parameters (such as rates of leaf turnover, respiration, photosynthesis 
per leaf area) in the model are linked via assumed tradeoffs. For
example, the trait LMA is used to estimate the rate of leaf turnover,
based on widely observed scaling relationship from \citet{Wright-2004}.

@@

With above functionality, \texttt{TREE} can be used to estimate essential
physiological rates for individual plants (Fig. \ref{fig:plant}). The
function \texttt{grow\_plant\_to\_size} takes a given strategy and light
environment and grows the plant, producing a trajectory of size over
time (Fig. \ref{fig:plant}a). This is achieved by integrating an
Ordinary Differential Equation (ODE) of size-dependent growth rate (see
Appendix \ref{sec:demography} for more
details). As the plant grows, allocation to different tissue types varies, and
the composition of the plant changes to include more stem and less
leaf (Fig. \ref{fig:plant}b). This in turn affects maintenance and
turnover costs and the growth rate of plants (Fig. \ref{fig:plant}c).
This shift in allocation alone generates a distinctive hump-shaped pattern of absolute
growth with size \citep{King-2011}, and equally the widely regarded
decline in relative growth rate with size from birth onwards
\citep{Enquist-2007}.

By varying the light environment and measuring growth rate, the whole
plant light compensation point (WPLCP) can be computed (Fig.
\ref{fig:plant}d). The WPLCP is the light level at which the plant 
stops growing, and is increasingly regarded as the most useful
measure of a strategy's shade tolerance
\citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, WPLCP
increases with plant size, due to increased costs of building and
maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, WPLCP
decreases with LMA, because high LMA species have lower leaf turnover
\citep{Baltzer-2007, Lusk-2013}.

@@

When modelling a patch, we are interested in modelling changes in the
size-density-distribution \(n(h,a)\) over time, i.e.~the density of
individuals with height \(h\) within a patch of age \(a\). We assume
that patches are large, and are vertically but not horizontally
structured. Thus, shading in \texttt{TREE} is modelled as a mean field with
respect to spatial layout. Similar assumptions are found in many models
simulating size-structured dynamics
\citep{Moorcroft-2001, Huston-1987, Smith-2014}. Under these
assumptions, the dynamics of \(n\) behave deterministically and can be
approximated via a Partial Differential Equation (PDE) (see Appendix
\ref{sec:demography}). The same PDE has been shown to capture average behaviour
across a large number of small patches \citep{Moorcroft-2001}. Ideally
one would also consider spatial interactions within patches,
however such models are very computationally demanding, and also
impossible to solve in a deterministic fashion \citep{Pacala-1996}.
Moreover, it remains unclear whether resolving spatial details within a
patch would provide further ecological detail, in part because the
process of competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

Our approach for modelling size-structured population dynamics
is based on the Escalator Boxcar Train technique (EBT)
\citep{Deroos-1997, Deroos-1992, Deroos-1988}. The EBT solves the PDE
describing development of \(n\) by approximating the density function
with a collection of cohorts spanning the size spectrum. Following a
disturbance, a series of cohorts are introduced into each patch. These
cohorts are then transported along the characteristic equations of the
PDE. Biologically these are the growth trajectories of individuals,
conditioned on them surviving (e.g. Fig. \ref{fig:patch}a). Meanwhile,
growth and mortality combine to alter the density of individuals along
the trajectories (Fig. \ref{fig:patch}b).

@@

The extension for handling strong size-asymmetric competition involves
adaptive refining of time points at which new cohorts are introduced
into the population. In the original EBT, cohorts were introduced at a
fixed time interval. Under competition, the growth trajectories of
individuals born at similar times can diverge substantially over time
(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography}).
This is can become a problem later during patch
development because much of the population can be located between
widely-spaced cohorts, delivering low numerical precision. To maintain
numerical precision we therefore use an iterative algorithm to adaptively refine
the cohort spacing until the trajectories of adjacent cohorts remain
adequately resolved throughout development of the patch (see
Appendix \ref{sec:demography} for details). This refinement causes an uneven
distribution of cohort introduction times across patch age (Fig.
\ref{fig:patch}a), generally with tighter spacing of cohorts in 
younger patches.

@@

First, \texttt{TREE} can simulate patterns of patch development after
disturbance, including self-thinning behaviours and successional
replacement. Whereas previous approaches for modelling self-thinning
were limited to a single species
\citep[e.g.][]{Barnes-2004, Coomes-2007}, \texttt{TREE} accommodates multiple
 species with different traits (Fig. \ref{fig:schematic}b; Fig.
\ref{fig:patch}).

@@

Having simulated a single patch, it is straightforward to scale up from
a single patch to an entire metapopulation, at least for the situation
of metapopulations at an equilibrium mix of patch ages and where
disturbances remove all established vegetation. The frequency-density
\(p(a)\) of patches aged \(a\) in the landscape is needed. Assuming there
are a large number of patches means the dynamics of \(p\) behave
deterministically and can be approximated via a second PDE
\citep{Vonfoerster-1959, Mckendrick-1926} (see Appendix
\ref{sec:demography} for
details). Scaling from patch to metapopulation is then achieved by
weighting the temporal dynamics of a single patch by \(p(a)\), the
relative abundance of patches aged \(a\) in the metapopulation.

The main numerical challenge in stepping from single patch
to metapopulation is to identify the equilibrium seed rain of the
metapopulation. For a given seed rain input, the model produces a seed rain output. 
The equilibrium seed rain is the vector of seed rain
where input seed rain equals output seed rain. Finding this equilibrium
requires finding a root of a multidimensional function, subject to the
constraints that this root is stable (e.g., the trivial equilibrium of
zero seed rain always exists, but is often not stable) (Appendix \ref{sec:fitness});

@@

Beyond its conceptual appeal, this idea of a dynamic equilibrium allows
us to close the demographic feedback loop and thereby create a
self-regulating system. The equilibrium seed rain for multi-species
communities is approached demographically, incorporating the combined
effects of the model's physiological rules, species traits, competitive
interactions and the prevailing disturbance regime on the
size-distributions of plants within a metapopulation of patches. Unlike
some vegetation models, there is no need to set the density of seeds or
seedlings, these arise as outputs from the model. Characteristics of the
vegetation such as total leaf area, biomass, productivity and allocation
to different plant parts also arise as emergent properties (Fig.
\ref{fig:patch}c). Similarly, distributions of abundance, leaf area and
growth rate with respect to size emerge at both patch scale and scale of
metapopulation (Fig. \ref{fig:emergent}). Note in particular how a single
physiological model produces a variety of relationships, dependent on patch
age. These emergent phenomena might then be compared to the observed
distributions in large forest tracts \citep{Muller-2006}.

@@

With the extension to metapopulation, we are modelling demography across
the entire life cycle and are thus in a position to estimate the fitness
of different types. Invasion fitness quantifies the ability of a rare
individual with particular traits to establish in a community of
established residents at their equilibrium densities \citep{Metz-1992}.
Invasion fitness
is ideally calculated as the long-term per capita growth rate of the new
type; however, in structured metapopulations, a more convenient
indicator of per-capita growth rate is given by the logarithm of the
basic reproduction ratio \citep{Gyllenberg-2001, Metz-2001}. The basic
reproduction ratio \(R\) is simply the total seed output of the new
type, averaged across the metapopulation. The new type can invade when
\(R > 1\).

@@

The current parametrisation of the physiological model is based on
\citet{Falster-2011}. Parameter values were drawn from available literature (see
Appendix \ref{sec:FFW16}); however for some parameters limited or no data were
readily available. Providing updated parameter sets is trivial (see Appendix
\ref{sec:parameters}). Ideally, parameters would be inferred through a combination of
independent estimates and constraining outputs from \texttt{TREE} against diverse data sources
\citep{Lebauer-2012, Keenan-2013}.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/schematic}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/plant.pdf}
\caption{Physiological dynamics for individual plants vary with size, trait
and light environment. (a) Growth trajectories for plants differing in light
environment and trait (LMA, low = 0.1, solid line; high =  0.267, dashed
line). Lower light environments slow trajectories and interact non-linearly
with traits. (b) Over time, the fraction of living tissue switches from leaf
towards sapwood, with high LMA species having relatively more mass in leaf
than low LMA species. (c) Growth rates peak around 5m high, but this position
varies with both trait and light level (note that this is the derivative of
panel (a)). (d) Growth rates vary with plant size (seedling and mature shown)
and with traits; growth rate differences are more pronounced in high light.
Whole plant light compensation points emerge from the model as the point where
growth rate is zero (x intercept).  The light level used in panels (a) and (c)
is indicated with dots.} [Could you put delta symbols on the y axis rather than d? Also there is alot going in these figures with 4 lines each. You could make it easier for the reader by adding the dashed lines to the keys]
\label{fig:plant}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/patch.pdf}
\caption{Population dynamics for two species competing within a patch. Lines
in panels (a) same growth trajectories for individuals on the
boundary of successive "cohorts" for each species, using a schedule of 
introduction times generated by an adaptive refinement algorithm.
 Red lines are the low LMA
species from figure \ref{fig:plant}, blue lines are the high LMA species.  The
initial growth rate advantage of the low LMA species (see Figure
\ref{fig:plant} c) means it quickly over tops the high LMA species,
suppressing its growth.  Self-thinning leaves space for the high LMA species
to establish below the canopy of the fast species. Panel (b) shows the same
trajectories as in panel (a), but with shading indicating the density $n$ of
individuals at given size and patch age (light = low density, dark = high density).  
(c) The leaf area index at ground
level for each species.  The dotted line is the total leaf area index (sum
across both species).}
\label{fig:patch}
\end{figure}
